---
title: 'Analysis: Exosymbiont abundances'
output: html_document
date: "2025-05-26"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r clear environment}
# clear environment for maximum reproducibility
rm(list=ls())
```

```{r load packages}
# load librarian (package for installing/loading packages)
if (!require("librarian")) install.packages("librarian")

# load requried packages
librarian::shelf(here, # relative file paths 
                 tidyverse, # data wrangling
                 BlakeRMills/MoMAColors, # color palette
                 car, # type II Anova for comparing model terms
                 glmmTMB, # modeling
                 DHARMa, # for residual exploration: normality of residuals, dispersion, etc.
                 patchwork, # for putting figures together
                 performance, # for check homogeneity
                 emmeans # for contrasts
                 )
```

```{r explicitly assign functions if needed}
# make sure correct functions attached to the correct packages
summarize <- dplyr::summarize
group_by <- dplyr::group_by
select <- dplyr::select
```

```{r set seed}
set.seed(0429)
```

# Set the color palette and graphics

```{r consistent color palette by treatment}
colorz <- moma.colors("Warhol", n=15, type="discrete")
b.color <- colorz[3]
control.color <- colorz[9]
crab.color <- colorz[1]
n.color <- colorz[6]
```

```{r set graphing theme}
# ggplot theme
theme_set(
  theme_bw(base_size = 20) +
    theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
    theme(text = element_text(size=20)) 
) 
```

# Read in data

```{r load data}
# full exosymbiont count data
exosym <- read_csv(here("data", "exosymbionts.csv")) %>% 
  # exclude coral that experienced 100% tissue loss
  filter(coral_id != 25)

# metadata (treatments, block, coral volume, etc.)
metadata <- read_csv(here("data", "metadata.csv")) %>% 
  # exclude coral that experienced 100% tissue loss
  filter(coral_id != 25) %>% 
  # get column names/values nice for plotting
  mutate(Nutrients = case_when(nutrients == "Y" ~ "Enriched",
                               nutrients == "N" ~ "Ambient")) %>% 
  mutate(Crab = case_when(crab == "Y" ~ "Crab",
                               crab == "N" ~ "No crab"))
```

# Volume differences

We want to: 
1. Make sure there are no differences in volume among treatments 
2. Examine the effect of coral volume on total exosymbiont abundance

For that, first we need to calculate the total number of exosymbionts per colony and then join that information with our metadata on coral volume 

```{r get total macrosymbionts per colony and join metadata}
exosym %>% 
  group_by(coral_id) %>% 
  summarize(abundance = sum(abundance)) %>% 
  full_join(metadata) %>% 
  mutate(density_cm3 = abundance/coral_vol_cm3,
         density_1000cm3 = 1000*abundance/coral_vol_cm3) -> total_abund
```

## Model

```{r test volume differences}
m.vol <- glmmTMB(coral_vol_cm3 ~ treatment + (1|block), 
                    family = Gamma(link = "log"), 
                    data = total_abund) 
```

```{r test assumptions}
# get residuals
resid.m.vol <- DHARMa::simulateResiduals(m.vol)

testResiduals(resid.m.vol, plot = T)
testQuantiles(resid.m.vol, plot = T)
check_homogeneity(m.vol)
check_model(m.vol)
```

```{r examine results}
car::Anova(m.vol, type = 2) 
```

# Macrosymbiont densities

First, some overall summary statistics for the results section

```{r range of macrosymbionts per coral}
# total number of macrosymbionts counted, mean, and standard deviation
total_abund %>% 
  summarize(total_abundance = sum(abundance),
            mean_abundance = mean(abundance),
            sd_abundance = sd(abundance),
            N = n(),
            se_abundance = sd_abundance/sqrt(N)
            ) 
```

Visually examine differences in density

```{r plot total abundance differences}
pdf(here("figures", "total_macrosymbiont_density.pdf"), height = 5, width = 9)

total_abund %>% 
  # get column names/values nice for plotting
  mutate(Nutrients = case_when(nutrients == "Y" ~ "Enriched",
                               nutrients == "N" ~ "Ambient")) %>% 
  mutate(Crab = case_when(crab == "Y" ~ "Crab",
                               crab == "N" ~ "No crab")) %>% 
  # plot
  ggplot(aes(x = Nutrients, y = density_1000cm3, fill = Crab)) +
  geom_boxplot(size = 1,
               outlier.shape = NA) + 
  geom_point(size = 2.2,
             color = "black",
             position = position_jitterdodge(jitter.width = 0.1,
                                             jitter.height = 0,
                                             dodge.width = 0.75)) +
  scale_fill_manual(values = c("No crab" = control.color, "Crab" = crab.color)) +
  theme(text = element_text(size=20),
        panel.border = element_rect(color="black", size=2.3)) +
  ylab(expression(paste("Exosymbiont abundance/1000", cm^3))) +
  xlab("")

dev.off()
```

## Model

```{r tweedie}
m.abund.1 <- glmmTMB(density_1000cm3 ~ crab + nutrients + crab:nutrients + (1|block), 
                           family = tweedie(link = "log"),
                           data = total_abund)
```

```{r test assumptions}
# get residuals
resid.m.abund.1 <- DHARMa::simulateResiduals(m.abund.1)

testResiduals(resid.m.abund.1, plot = T)
testDispersion(resid.m.abund.1)
testZeroInflation(resid.m.abund.1)
testQuantiles(resid.m.abund.1, plot = T) 
check_model(m.abund.1)
```

```{r examine results}
car::Anova(m.abund.1, type = 3, contrasts = list(nutrients = contr.sum, crab = contr.sum))
```

Drop interaction

```{r tweedie}
m.abund.2 <- glmmTMB(density_1000cm3 ~ crab + nutrients + (1|block), 
                           family = tweedie(link = "log"),
                           data = total_abund)
```

```{r test assumptions}
# get residuals
resid.m.abund.2 <- DHARMa::simulateResiduals(m.abund.2)

testResiduals(resid.m.abund.2, plot = T)
testDispersion(resid.m.abund.2)
testZeroInflation(resid.m.abund.2)
testQuantiles(resid.m.abund.2, plot = T) 
check_model(m.abund.2)
```

```{r examine results}
car::Anova(m.abund.2, type = 2)
```


# A mutualist: D. flavicaudus

## Model

```{r D. flav ~ treatments}
m.dflav.1 <- glmmTMB(dflav_density_per_1000cm3 ~ crab + nutrients + crab:nutrients + (1|block), 
                     family =  tweedie(link = "log"),
                     data = metadata)
```

```{r test assumptions}
# get residuals
resid.dflav.1 <- DHARMa::simulateResiduals(m.dflav.1)

testResiduals(resid.dflav.1, plot = T)
testDispersion(resid.dflav.1)
testZeroInflation(resid.dflav.1)
testQuantiles(resid.dflav.1, plot = T) 
check_model(m.dflav.1)
```

```{r examine results}
car::Anova(m.dflav.1, type = 3, contrasts = list(nutrients = contr.sum, crab = contr.sum))
```

Drop the interaction

```{r D. flav ~ treatments}
m.dflav.2 <- glmmTMB(dflav_density_per_1000cm3 ~ crab + nutrients +  (1|block), 
                     family =  tweedie(link = "log"),
                     data = metadata)
```

```{r test assumptions}
# get residuals
resid.dflav.2 <- DHARMa::simulateResiduals(m.dflav.2)

testResiduals(resid.dflav.2, plot = T)
testDispersion(resid.dflav.2)
testZeroInflation(resid.dflav.2)
testQuantiles(resid.dflav.2, plot = T) 
check_model(m.dflav.2)
```

```{r examine results}
car::Anova(m.dflav.2, type = 2)
```

Get magnitude of nutrient effect:

```{r}
summary(emmeans::emmeans(m.dflav.2, pairwise ~ nutrients), type = "response")$emmeans 
```

## Plot

First, get contrasts for nutrients, which was a significant predictor of variation:

```{r}
emm.dflav <- as_tibble(summary(emmeans::emmeans(m.dflav.2, pairwise ~ nutrients), type = "response")$emmeans) 
```


```{r}
emm.dflav %>% 
  # get column names/values nice for plotting
  mutate(Nutrients = case_when(nutrients == "Y" ~ "Enriched",
                               nutrients == "N" ~ "Ambient")) %>% 
  # just for plotting purposees:
  mutate(Crab = "No crab") %>% 
  ggplot() +
  geom_pointrange(aes(x = Nutrients, y = response,
                      ymin = asymp.LCL, ymax = asymp.UCL), 
                  size = 1.5, linewidth = 1.5, color = "black") +
  # make a new color scale
  ggnewscale::new_scale_color() +
  geom_jitter(data = metadata,
              aes(x = Nutrients, y = dflav_density_per_1000cm3, color = Crab), 
              height = 0, width = 0.1, size = 2) +
  scale_color_manual(values = c("No crab" = control.color, "Crab" = crab.color), 
                     name = "") +
  ylab(expression(paste("Predicted abundance/1000", cm^3))) +
  xlab("Nutrient treatment") +
  theme(legend.position = "top",
        panel.border = element_rect(color="black", size=2.3)) +
  xlab("") -> plt.dflav

```

# A parasite: G. monodonta

## Model

The Tweedie model has convergence problems, which is why we are using a Gamma family error distribution

```{r}
m.gmono.1 <- glmmTMB(gmono_density_per_1000cm3 ~ crab + nutrients + crab:nutrients + (1|block), 
                     family = Gamma(link = "log"), 
                     data = metadata) 
```

```{r test assumptions}
# get residuals
resid.gmono.1 <- DHARMa::simulateResiduals(m.gmono.1)

testResiduals(resid.gmono.1, plot = T)
testDispersion(resid.gmono.1)
testZeroInflation(resid.gmono.1)
testQuantiles(resid.gmono.1, plot = T) 
check_model(m.gmono.1)
```

Looks all right. Check results

```{r examine results}
car::Anova(m.gmono.1, type = 3, contrasts = list(nutrients = contr.sum, crab = contr.sum))
```

Drop interaction

```{r}
m.gmono.2 <- glmmTMB(gmono_density_per_1000cm3 ~ crab + nutrients + (1|block), 
                     family = Gamma(link = "log"), 
                     data = metadata) 
```

```{r test assumptions}
# get residuals
resid.gmono.2 <- DHARMa::simulateResiduals(m.gmono.2)

testResiduals(resid.gmono.2, plot = T)
testDispersion(resid.gmono.2)
testZeroInflation(resid.gmono.2)
testQuantiles(resid.gmono.2, plot = T) 
check_model(m.gmono.2)
```

Looks all right. Check results

```{r examine results}
car::Anova(m.gmono.2, type = 2)
```

Get magnitude of nutrient effect:

```{r}
summary(emmeans::emmeans(m.gmono.2, pairwise ~ crab), type = "response")$emmeans 
```

## Plot

Get contrasts using emmeans

```{r}
emm.gmono <- as_tibble(summary(emmeans::emmeans(m.gmono.2, pairwise ~ crab), type = "response")$emmeans) 
```


Plot with raw data:

```{r}
emm.gmono %>%
  # get column names/values nice for plotting
  mutate(Crab = case_when(crab == "Y" ~ "Crab",
                               crab == "N" ~ "No crab")) %>% 
  # make a fake nutrients variable for plotting
  mutate(Nutrients = "Enriched") %>% 
  ggplot() +
  geom_pointrange(aes(x = Crab, y = response,
                      ymin = asymp.LCL, ymax = asymp.UCL), 
                  color = "black", size = 1.5, linewidth = 1.5) +
  # make a new color scale
  ggnewscale::new_scale_color() +
  geom_jitter(data = metadata,
              aes(x = Crab, y = gmono_density_per_1000cm3, color = Nutrients),
              height = 0, width = 0.1, size = 2) +
  scale_color_manual(values = c("Ambient" = b.color, "Enriched" = n.color), 
                     name = "") +
  ylab(expression(paste("Predicted abundance/1000", cm^3))) +
  xlab ("Crab treatment") +
  theme(legend.position = "top",
        panel.border = element_rect(color="black", size=2.3)) +
  xlab("") -> plt.gmono
```


# Plot together

```{r}
pdf(here("figures", "parasite_mutualist_predictions.pdf"), height = 5.5, width = 10.5)

plt.gmono + plt.dflav + plot_annotation(tag_levels = c('a'), tag_prefix = '(',
                            tag_suffix = ')')

dev.off()
```



