---
title: 'Analysis: Exosymbiont communities'
output: html_document
date: "2025-05-25"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r clear environment}
# clear environment for maximum reproducibility
rm(list=ls())
```

```{r load packages}
# load librarian (package for installing/loading packages)
if (!require("librarian")) install.packages("librarian")

# load requried packages
librarian::shelf(here, # relative file paths 
                 tidyverse, # data wrangling
                 BlakeRMills/MoMAColors, # color palette
                 vegan, # community/diversity analyses
                 patchwork, # joining figures together
                 glmmTMB, # linear models
                 car, # type II ANOVA
                 performance, # for residual exploration: normality of residuals, dispersion, etc.
                 DHARMa # for residual/model fit exploration
                 )
```

```{r explicitly assign functions if needed}
# make sure correct functions attached to the correct packages
summarize <- dplyr::summarize
group_by <- dplyr::group_by
select <- dplyr::select
```

```{r set seed}
set.seed(0429)
```

# Set the color palette and graphics

```{r consistent color palette by treatment}
colorz <- moma.colors("Warhol", n=15, type="discrete")
b.color <- colorz[3]
control.color <- colorz[9]
crab.color <- colorz[1]
n.color <- colorz[6]
```

```{r set graphing theme}
# ggplot theme
theme_set(
  theme_bw(base_size = 20) +
    theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
    theme(plot.title = element_text(vjust = -6))
) 
```

# Read in data

```{r}
exosym <- read_csv(here("data", "exosymbionts.csv")) %>% 
  # exclude the coral that experienced 100% tissue loss
  filter(coral_id != 25)

metadata <- read_csv(here("data", "metadata.csv")) %>% 
  filter(coral_id != 25)
```

# Get overall taxa abundances

For supplemental table describing taxa groups

```{r}
exosym %>% 
  group_by(taxa_group) %>% 
  summarize(abundance = sum(abundance)) %>% 
  arrange(-abundance) %>% 
  write_csv(here("tables", "supplemental_taxa_table.csv"))
```

# Create a taxa matrix

Want to pivot this into long format and fill NA's with zeros where no taxa groups were observed

```{r}
exosym %>% 
  pivot_wider(names_from = taxa_group,
              values_from = abundance,
              values_fill = 0) -> taxa_mat_1
```

# Calculate diversity metrics

Make a dataframe for diversity metrics

```{r}
taxa_diversity_1 <- taxa_mat_1
```

```{r use vegan functions}
taxa_diversity_1$richness <- specnumber(taxa_mat_1[,-1])
taxa_diversity_1$shannon <- diversity(taxa_mat_1[,-1], index = "shannon")
taxa_diversity_1$evenness <- taxa_diversity_1$shannon/log(taxa_diversity_1$richness)
```

Add in metadata

```{r}
taxa_diversity_1 %>% 
  inner_join(metadata) %>% 
  # get column names/values nice for plotting
  mutate(Crab = case_when(crab == "Y" ~ "Crab",
                               crab == "N" ~ "No crab")) %>%
  mutate(Nutrients = case_when(nutrients == "Y" ~ "Enriched",
                               nutrients == "N" ~ "Ambient")) -> taxa_diversity_2
```

# Examine richness differences

Richness plot

```{r plot}
taxa_diversity_2 %>% 
  ggplot(aes(x = Nutrients, y = richness, fill = Crab)) +
  geom_boxplot(size = 1,
               outlier.shape = NA) + 
  geom_point(size = 2.2,
             color = "black",
             position = position_jitterdodge(jitter.width = 0.1,
                                             jitter.height = 0,
                                             dodge.width = 0.75)) +
  ylab("Richness") +
  scale_fill_manual(values = c("No crab" = control.color, "Crab" = crab.color)) +
  theme(text = element_text(size=20),
        panel.border = element_rect(color="black", fill = NA, size = 2.3),
        legend.position = "none") -> rich.plt
```

## Model

```{r model}
m.rich.1 <- glmmTMB(richness ~ crab + nutrients + crab:nutrients + (1|block),
                    family = poisson(),
                    data = taxa_diversity_2)
```

```{r test assumptions}
resid.m.rich.1 <- DHARMa::simulateResiduals(m.rich.1)

testResiduals(resid.m.rich.1, plot = T)
testDispersion(resid.m.rich.1)
testZeroInflation(resid.m.rich.1)
testQuantiles(resid.m.rich.1, plot = T) 
check_model(m.rich.1)
```

Examine results

```{r examine results}
car::Anova(m.rich.1) 
```

# Evenness

```{r plot}
taxa_diversity_2 %>% 
  ggplot(aes(x = Nutrients, y = evenness, fill = Crab)) +
  geom_boxplot(size = 1,
               outlier.shape = NA) + 
  geom_point(size = 2.2,
             color = "black",
             position = position_jitterdodge(jitter.width = 0.1,
                                             jitter.height = 0,
                                             dodge.width = 0.75)) +
  scale_fill_manual(values = c("No crab" = control.color, "Crab" = crab.color)) +
  ylab("Evenness") + 
  theme(text = element_text(size=20),
        panel.border = element_rect(color="black", size=2.3),
        legend.position = "none") -> even.plt
```

## Model

```{r first model}
m.even.1 <- glmmTMB((evenness) ~ crab + nutrients + crab:nutrients  + (1|block),
                  family = Gamma(link = "log"),
                  data = taxa_diversity_2)
```

```{r test assumptions}
resid.m.even.1 <- DHARMa::simulateResiduals(m.even.1)

testResiduals(resid.m.even.1, plot = T)
testDispersion(resid.m.even.1)
testZeroInflation(resid.m.even.1)
testQuantiles(resid.m.even.1, plot = T) 
check_model(m.even.1)
```

```{r examine results}
car::Anova(m.even.1) 
```

# Beta diversity

Calculate Bray Curtis distances

```{r calculate distances}
bray_mat <- vegdist(taxa_mat_1[,-1], "bray") # 1 is maximum dissimilarity
```

Make sure key aligns

```{r}
unique(taxa_mat_1[,1] == metadata[,1])
```

Calculate dispersion 

```{r calculate beta dispersion}
betaDisper<- vegan::betadisper(bray_mat, metadata$treatment) 

# are groups a good predictor of dispersion?
anova(betaDisper) # no

# plot of dispersion test
plot(betaDisper, sub = "") 
boxplot(betaDisper, main = "Treatment", xlab = "") 
```

Plot by hand. Get distances:

```{r}
metadata$distance_centriod <- betaDisper$distances
```

Plot

```{r}
metadata %>% 
  mutate(Crab = case_when(crab == "Y" ~ "Crab",
                               crab == "N" ~ "No crab")) %>%
  mutate(Nutrients = case_when(nutrients == "Y" ~ "Enriched",
                               nutrients == "N" ~ "Ambient")) %>% 
  ggplot(aes(x = Nutrients, y = distance_centriod, fill = Crab)) +
  geom_boxplot(size = 1,
               outlier.shape = NA) + 
  geom_point(size = 2.2,
             color = "black",
             position = position_jitterdodge(jitter.width = 0.1,
                                             jitter.height = 0,
                                             dodge.width = 0.75)) +
  ylab("Distance to centroid") +
  scale_fill_manual(values = c("No crab" = control.color, "Crab" = crab.color)) + 
  theme(text = element_text(size=20),
        panel.border = element_rect(color="black", size=2.3)) -> beta.plt
```

## PERMANOVA

```{r}
adonis2(taxa_mat_1 ~ crab + nutrients + crab:nutrients, 
        strata = metadata$block, 
        data = metadata, 
        method = "bray",
        perm = 10000)
```

## Plot all together

```{r}
pdf(here("figures", "evenness_and_richness.pdf"), height = 9.5, width = 13)

(rich.plt + even.plt) / beta.plt + plot_annotation(tag_levels = c('a'), tag_prefix = '(',
                            tag_suffix = ')')

dev.off()
```