---
title: "Piecewise SEM protein and carb models"
output: html_document
date: "2025-05-26"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r clear environment}
# clear environment for maximum reproducibility
rm(list=ls())
```

```{r load packages}
# load librarian (package for installing/loading packages)
if (!require("librarian")) install.packages("librarian")

# load requried packages
librarian::shelf(here, # relative file paths 
                 tidyverse, # data wrangling
                 BlakeRMills/MoMAColors, # color palette
                 piecewiseSEM, # structural equation modeling
                 semEff, # indirect effect p-values
                 glmmTMB,
                 performance # check models
                 )
```

```{r explicitly assign functions if needed}
# make sure correct functions attached to the correct packages
summarize <- dplyr::summarize
group_by <- dplyr::group_by
select <- dplyr::select
```

```{r set seed}
set.seed(0429)
```

# Set the color palette and graphics

```{r consistent color palette by treatment}
colorz <- moma.colors("Warhol", n=15, type="discrete")

b.color <- colorz[3]
control.color <- colorz[9]
crab.color <- colorz[1]
n.color <- colorz[6]
```

```{r set graphing theme}
# ggplot theme
theme_set(
  theme_bw(base_size = 20) +
    theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
    theme(text = element_text(size=20)) 
) 
```

# Read in data

```{r}
physiology_1 <- read_csv(here("cleaned_data", "physiology.csv"))
metadata <- read_csv(here("cleaned_data", "metadata.csv"))
```

# Prepare data

Clean, join, and scale data

```{r}
metadata %>% 
  # join metadata and physiology data
  inner_join(physiology_1) %>% 
  # turn categories into indicators
  mutate(crab = case_when(
    crab == "Y" ~ 1,
    crab == "N" ~ 0,
    TRUE ~ -9999
    )) %>% 
  mutate(nutrients = case_when( 
    nutrients == "Y" ~ 1,
    nutrients == "N" ~ 0,
    TRUE ~ -9999
    )) %>% 
  mutate(interaction = nutrients*crab) %>% 
  # scale variables
  mutate(protein_change_scale = scale(protein_change)[,1],
         carb_change_scale = scale(carb_change)[,1],
         endosym_change_scale = scale(endosym_percent_change)[,1],
         dflav_density_scale = scale(dflav_density_per_1000cm3)[,1],
         gmono_density_scale = scale(gmono_density_per_1000cm3)[,1]
         ) -> scaled_data
```

# Make individual models

First we'll fit individual models. Most will be shared between protein and carb models

# Overall models

D. flavicaudus model

```{r D. flavicaudus model}
m.dflav <- glm(dflav_density_scale ~ nutrients,
               family = gaussian(link = "identity"),
               data = scaled_data)
```

```{r check model}
check_model(m.dflav)
  # check normality of residuals manually:
  qqnorm(m.dflav$residuals)
  qqline(m.dflav$residuals)
  check_normality(simulate_residuals(m.dflav))
```

G. monodonta model

```{r G. monodonta model}
m.gmono <- glm(gmono_density_scale ~ crab,
               family = gaussian(link = "identity"),
               data = scaled_data)
```

```{r check model}
check_model(m.gmono)
  # check normality of residuals manually:
  qqnorm(m.gmono$residuals)
  qqline(m.gmono$residuals)
  check_normality(simulate_residuals(m.gmono))
```

Endosymbiont density model

```{r endosymbiont model}
m.endo <- glm(endosym_change_scale ~ nutrients,
              family = gaussian(link = "identity"),
              data = scaled_data)
```

```{r check model}
check_model(m.endo)
  # check normality of residuals manually:
  qqnorm(m.endo$residuals)
  qqline(m.endo$residuals)
  check_normality(simulate_residuals(m.endo))
```

# Protein model

```{r protein model}
m.prot.health <- glm(protein_change_scale ~ nutrients + crab + interaction + endosym_change_scale + dflav_density_scale + gmono_density_scale,
                     family = gaussian(link = "identity"),
                     data = scaled_data)
```

```{r check model}
check_model(m.prot.health)
  qqnorm(m.prot.health$residuals)
  qqline(m.prot.health$residuals)
```

Create pSEM

```{r pSEM}
sem.prot <- psem(
  m.dflav, 
  m.gmono, 
  m.endo, 
  m.prot.health,
  data = as.data.frame(scaled_data)
)
```

```{r summary}
summary(sem.prot, conserve = TRUE)
```

**Notes on the summary output:**
- For tests of directed separation, p-values > 0.05 means those paths are justified to exclude
- Fischer's C statistic > 0.05 provides support for the hypothesized model structure

## Estimate direct and indirect effects

Use the `semEff` package to calculate effects via bootstrapping

```{r bootstrap effects}
# bootstrap and save standardized effects
protein.sem.boot <- bootEff(sem.prot, R = 20000, 
                            seed = 0429,
                            parallel = "multicore") 

# calculate direct, indirect, and total effects
protein.sem.eff <- semEff(protein.sem.boot)
```

```{r examine paths}
summary(protein.sem.eff, 
        ci.conf = 0.95) # defaults to 95% CI and standardized, but just being thorough 
```

**Results**

*Protein change*

                                   Effect     Bias   Std. Err.   Lower CI Upper CI    
                                   ------   ------   ---------   -------- --------    
 DIRECT:    nutrients            | -0.347 |  0.017 |     0.131 |   -0.595   -0.092 | *
            crab                 | -0.232 |  0.011 |     0.146 |   -0.485    0.088 |  
            interaction          |  0.144 | -0.013 |     0.148 |   -0.168    0.410 |  
            dflav.density.scale  |  0.449 | -0.020 |     0.133 |    0.171    0.675 | *
            gmono.density.scale  | -0.253 |  0.026 |     0.140 |   -0.545   -0.002 | *
            endosym.change.scale |  0.154 | -0.011 |     0.117 |   -0.052    0.411 |  
                                 |        |        |           |                   |  
 INDIRECT:  nutrients            |  0.298 | -0.016 |     0.086 |    0.148    0.490 | *
            crab                 |  0.085 | -0.007 |     0.056 |    0.001    0.240 | *
                                 |        |        |           |                   |  
 TOTAL:     nutrients            | -0.049 |  0.001 |     0.130 |   -0.333    0.184 |  
            crab                 | -0.147 |  0.003 |     0.156 |   -0.426    0.193 |  
            interaction          |  0.144 | -0.013 |     0.148 |   -0.168    0.410 |  
            dflav.density.scale  |  0.449 | -0.020 |     0.133 |    0.171    0.675 | *
            gmono.density.scale  | -0.253 |  0.026 |     0.140 |   -0.545   -0.002 | *
            endosym.change.scale |  0.154 | -0.011 |     0.117 |   -0.052    0.411 |  
                                 |        |        |           |                   |  
 MEDIATORS: dflav.density.scale  |  0.246 | -0.016 |     0.077 |    0.112    0.410 | *
            gmono.density.scale  |  0.085 | -0.007 |     0.056 |    0.001    0.240 | *
            endosym.change.scale |  0.052 |  0.000 |     0.050 |   -0.009    0.203 |  

*D. flavicaudus*

                        Effect     Bias   Std. Err.   Lower CI Upper CI    
                        ------   ------   ---------   -------- --------    
 DIRECT:    nutrients |  0.548 | -0.002 |     0.106 |    0.284    0.715 | *

*G. monodonta*

                   Effect     Bias   Std. Err.   Lower CI Upper CI    
                   ------   ------   ---------   -------- --------    
 DIRECT:    crab | -0.337 | -0.004 |     0.115 |   -0.524   -0.048 | *
 
*Endosymbionts*
 
                        Effect     Bias   Std. Err.   Lower CI Upper CI    
                        ------   ------   ---------   -------- --------    
 DIRECT:    nutrients |  0.336 | -0.001 |     0.139 |    0.008    0.566 | *


# Carbohydrate model

```{r make carbohydrate model}
m.carb.health <- glm(carb_change_scale ~ nutrients + crab + interaction + endosym_change_scale + dflav_density_scale + gmono_density_scale,
                     data = scaled_data)
```

```{r examine results}
check_model(m.carb.health)
  qqnorm(m.carb.health$residuals)
  qqline(m.carb.health$residuals)
```

```{r build pSEM}
sem.carb <- psem(
  m.dflav, 
  m.gmono, 
  m.endo, 
  m.carb.health,
  data = as.data.frame(scaled_data)
)
```

```{r examine fit}
summary(sem.carb, conserve = TRUE)
```

## Estimate direct and indirect effects

Use the `semEff` package to calculate effects via bootstrapping

```{r bootstrap effects}
# bootstrap and save standardized effects
carb.sem.boot <- bootEff(sem.carb, R = 20000, 
                         seed = 0429) 

# calculate direct, indirect, and total effects
carb.sem.eff <- semEff(carb.sem.boot)
```

```{r examine paths}
summary(carb.sem.eff, 
        ci.conf = 0.95) # defaults to 95% CI and standardized, but just being thorough 
```

**Results**

*Carbohydrate change*

                                   Effect     Bias   Std. Err.   Lower CI Upper CI  
                                   ------   ------   ---------   -------- --------  
 DIRECT:    nutrients            | -0.125 |  0.004 |     0.130 |   -0.375    0.133 |
            crab                 | -0.094 |  0.000 |     0.173 |   -0.398    0.291 |
            interaction          |  0.070 | -0.006 |     0.166 |   -0.262    0.389 |
            dflav.density.scale  |  0.111 | -0.002 |     0.153 |   -0.237    0.371 |
            gmono.density.scale  | -0.172 |  0.004 |     0.100 |   -0.366    0.041 |
            endosym.change.scale |  0.265 | -0.018 |     0.152 |   -0.025    0.567 |
                                 |        |        |           |                   |
 INDIRECT:  nutrients            |  0.150 | -0.008 |     0.108 |   -0.074    0.355 |
            crab                 |  0.058 | -0.001 |     0.041 |   -0.007    0.160 |
                                 |        |        |           |                   |
 TOTAL:     nutrients            |  0.025 | -0.004 |     0.114 |   -0.197    0.250 |
            crab                 | -0.036 | -0.001 |     0.178 |   -0.358    0.348 |
            interaction          |  0.070 | -0.006 |     0.166 |   -0.262    0.389 |
            dflav.density.scale  |  0.111 | -0.002 |     0.153 |   -0.237    0.371 |
            gmono.density.scale  | -0.172 |  0.004 |     0.100 |   -0.366    0.041 |
            endosym.change.scale |  0.265 | -0.018 |     0.152 |   -0.025    0.567 |
                                 |        |        |           |                   |
 MEDIATORS: dflav.density.scale  |  0.061 | -0.004 |     0.085 |   -0.129    0.212 |
            gmono.density.scale  |  0.058 | -0.001 |     0.041 |   -0.007    0.160 |
            endosym.change.scale |  0.089 | -0.004 |     0.065 |   -0.002    0.264 |
            
            
            